---
title: "Task 1"
author: "Matthieu Huy"
date: "2023-01-25"
output: html_document
---

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(readxl)
library(here)
library(lubridate)
library(cowplot)
```

### Read in data

This assignment uses amphibian abundance data recorded by the [Sierra Lakes Inventory Project](https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3)\

Citation:\
Knapp, R.A., C. Pavelka, E.E. Hegeman, and T.C. Smith. 2020. The Sierra Lakes Inventory Project: Non-Native fish and community composition of lakes and ponds in the Sierra Nevada, California ver 2. Environmental Data Initiative. https://doi.org/10.6073/pasta/d835832d7fd00d9e4466e44eea87fab3\ Accessed 1/30/2023

Metadata & info: Metadata is available [HERE](https://portal.edirepository.org/nis/metadataviewer?packageid=edi.577.2) and specifies species names. 

```{r}
amphibs <- read_excel(here("data/sierra_amphibians.xlsx"))
```

### Clean data for plot 1:\

- A graph of total mountain yellow-legged frog (Rana muscosa) count each year across all water bodies, by life stage excluding the ‘EggMass’ level.\

```{r, fig.width = 10, fig.height = 6}
date <- ymd(amphibs$survey_date[1:8150]) #create date column using lubridate
date <- lubridate::year(date) #pull just the year from ymd dates

amphibs$date = date #add new date column to data frame 

amphibs_plot1 <- amphibs |>
  select(date, species = amphibian_species, lifestage = amphibian_life_stage) |> #select cols
  filter(species == "RAMU" & lifestage != "EggMass") |> #select RAMU species & excl. Egg stage
  mutate_at(vars(1), as.character) |> #coerce dates column to character
  group_by(date, species) |> #group by date and species columns
  count(lifestage) |> #count # of observations for each lifestage each year
  ungroup()

```


### Plot 1

```{r, fig.width = 10, fig.height = 6}

plot1 <- ggplot() +
  geom_col(data = amphibs_plot1, 
           aes(x = date,
               y = n,
               fill = reorder(lifestage, n)), #color by lifestage, order lowest-highest freq
           position = position_dodge(width = 0.6), #set space between columns
           width = 0.6) + #set column width
  labs(x = "Year", #specify labels
       y = "Count",
       fill = "",
       caption = "Figure 1: Mountain yellow-legged frog population by year and life stage") +
  scale_x_discrete(breaks = c("1995", "1996", "1997", "2000", "2001", "2002")) + #create new scale
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10), #modify y scale to include more detail
                     limits = c(0, 500),
                     expand = c(0,0)) + #set limits of y
  scale_fill_manual(values = c("darkgoldenrod1", "darkgoldenrod3", "darkgoldenrod4"), #legend
                    labels = c("Tadpole", "Sub-Adult", "Adult")) +
  theme_light() +
  theme(axis.text = element_text(color = "black", size = 8),                    #fonts/font size
        axis.title.x = element_text(color = "black", size = 10, face = "bold"),  
        axis.title.y = element_text(color = "black", size = 10, face = "bold",
                                    vjust = 2), 
        plot.caption = element_text(hjust = 1.8, color = "black",
                                       size = 12, face = "bold"), 
        legend.position = "top",
        legend.text = element_text(size = 8))
```

### Clean data for plot 2:

- A column graph containing total counts (over all years of the study) of combined adult and subadult endangered mountain yellow-legged frogs (Rana muscosa) observed in the 5 lakes with the greatest total observed counts.

```{r}
amphibs_plot2 <- amphibs |>
  select(lake_id, species = amphibian_species, lifestage = amphibian_life_stage) |> #select cols
  filter(lifestage %in% c("Adult", "SubAdult") & species == "RAMU") |> #select only adult/sub of RAMU
  group_by(lake_id) |> #group by location observed
  count(lifestage) |> #count # of observations at each location for adults and sub-adults
  summarize_at("n", sum) |> #sum # of observations for both lifestages at each location
  ungroup() |>
  arrange(desc(n)) |> #arrange highest-lowest # of observations
  slice(1:15) |> #select top 5 greatest total counts (19, 12, 11, 10, 9), ties
  mutate_at("lake_id", as.character) #change lake_id col to character

species <- amphibs_plot1$species[1:15] #create vector for species (use for fill later)

amphibs_plot2$lake_id <- paste("Lake", amphibs_plot2$lake_id, sep = " ") #add "Lake" in front of ID#

amphibs_plot2 <- amphibs_plot2 |>
  mutate_at("lake_id", as.factor) |> #change lake_id to factor
  mutate(species = species) #add species column for ggplot fill purposes


```

### Plot 2

```{r, fig.width = 12, fig.height = 10}
plot2 <- ggplot() +
  geom_col(data = amphibs_plot2,
           aes(x = reorder(lake_id, n),
               y = n,
               fill = species), #color bars
           position = position_dodge(width = 0.25), #space between bars
           width = 0.5) + #width of bars
  coord_flip() +
  labs(x = "", #labels
       y = "Count",
       caption = "Figure 2: Greatest total observed populations across all years",
       fill = "") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5), #y axis breaks
                     limits = c(0, 20), #y axis range
                     expand = c(0,0)) + #remove gap between y axis and first category on x axis
  scale_fill_manual(values = c("darkgoldenrod2"), #color of bars
                    labels = c("")) +
  theme_light() +
  theme(axis.text = element_text(color = "black", size = 6),                    #fonts/font sizes
  axis.title.x = element_text(color = "black", size = 10, face = "bold"), 
  axis.title.y = element_text(color = "black", size = 10, face = "bold"), 
  plot.caption = element_text(hjust = -2.0, color = "black", 
                              size = 12, face = "bold"), 
  legend.title = element_text(size = 10),
  legend.position = "none", #remove legend
  panel.background = element_blank(),
  panel.grid.major = element_line(size = 0.25, linetype = 'solid', #other aesthetic stuff copied from lab
                                  colour = "lightgrey"),
  panel.grid.minor = element_line(size = 0.25, linetype = 'dotted',
                                  colour = "lightgrey"), 
  axis.ticks.x=element_blank(),
  axis.ticks.y=element_blank()) 
```


```{r, fig.width = 12, fig.height = 10}
plot_grid(plot1, plot2, labels = c("",""))
```


